---
title: "Compare two independent groups"
author: "Guillaume A. Rousselet"
date: "`r Sys.Date()`"
output:
  pdf_document:
    fig_caption: no
    number_sections: no
    toc: yes
    toc_depth: 2
    # github_document:
    # html_preview: yes
    # toc: yes
    # toc_depth: 2
---

# Dependencies
```{r, message=FALSE, warning=FALSE}
library(tibble)
library(ggplot2)
# library(cowplot)
source("./functions/theme_gar.txt")
source("./functions/Rallfun-v40.txt")
source("./functions/functions.txt")
```

```{r}
sessionInfo()
```

Wilcox's functions for two independent groups
```{r, eval = FALSE}
# to compare any estimators
pb2gen(x, y, alpha=0.05, nboot=1000, est=median, SEED=FALSE)
# same as pb2gen but only to compare medians:
medpb2()
# same as pb2gen but only to compare trimmed means:
trimpb2()
# to compare variances:
comvar2()
# For robust measures of scale, use pb2gen
```

# Illustrate populations

We sample from 2 populations that differ in skewness, leading to large differences in the tails.

```{r, fig.height=4}
meanlog1 <- 0
meanlog2 <- 0.2
sdlog1 <- 0.5
sdlog2 <- 0.7

x <- seq(0, 5, 0.1)

p <- ggplot(as.tibble(x), aes(x = x)) + theme_gar +
  stat_function(geom = "line", size = 1.5, aes(colour = "Lognormal(0, 0.5)"),
                fun = dlnorm, args = list(meanlog = meanlog1, sdlog = sdlog1)) +
  stat_function(geom = "line", size = 1.5, aes(colour = "Lognormal(0.2, 0.7)"),
                fun = dlnorm, args = list(meanlog = meanlog2, sdlog = sdlog2)) +
  labs(x = "Values", y = "Density") +
  scale_colour_manual(name = "", values = c("Lognormal(0, 0.5)" = "orange",
                                            "Lognormal(0.2, 0.7)" = "purple")) +
  theme(legend.position = c(.6, .8),
        axis.text.y=element_blank(),  #remove y axis labels
        axis.ticks.y=element_blank()) +
  guides(colour = guide_legend(override.aes = list(size = 3)))
p.pop <- p
p
```

## Population normalise difference

We use samples of 100,000 values to define the population normalise difference between 3rd quartiles.

```{r}
set.seed(21)
normdiff(q3(rlnorm(100000, meanlog = meanlog1, sdlog = sdlog1)), q3(rlnorm(100000, meanlog = meanlog2, sdlog = sdlog2)))
```

# Generate samples

We get 2 relatively large samples from 2 populations.

```{r}
set.seed(21)
n1 <- 50
n2 <- 60
samp1 <- rlnorm(n1, meanlog = meanlog1, sdlog = sdlog1)
samp2 <- rlnorm(n2, meanlog = meanlog2, sdlog = sdlog2)
```

Instead of the traditional measures of central tendency, like the mean or the median, here we're going to estimate the 3rd quartile of the marginal distributions. This can be justified by an interest in differences among slower responses. To illustrate that with the bootstrap we can build confidence intervals for any quantity, the group estimations will be on a normalised difference of quartiles: $effect = (q3_1-q3_2)/(q3_1+q3_2)$.

# Illustrate samples
```{r, fig.height=2}
set.seed(21) # for reproducible jitter
# raw data
df <- tibble(val = c(samp1, samp2),
             y = rep(1, n1+n2),
             gp = factor(c(rep("Group 1",n1),rep("Group 2",n2)))
             )

df.q2 <- tibble(y = rep(0.9,2), 
                yend = rep(1.1,2),
                x = c(q2(samp1),q2(samp2)), 
                xend = x,
                gp = factor(c("Group 1","Group 2"))
                )

df.q3 <- tibble(y = rep(0.9,2), 
                yend = rep(1.1,2),
                x = c(q3(samp1),q3(samp2)), 
                xend = x,
                gp = factor(c("Group 1","Group 2"))
                )

p <- ggplot(data = df, aes(x = val, y = y)) + theme_gar +
  # scatterplots
  geom_jitter(height = .05, alpha = 0.5, 
              size = 3, shape = 21, fill = "grey", colour = "black") +
  theme(axis.ticks.y = element_blank(),
    axis.text.y = element_blank(),
    axis.title.y = element_blank(),
    panel.grid.minor.x = element_blank()) +
  scale_y_continuous(breaks = 1) +
  # 3rd quartile
  geom_segment(data = df.q3, aes(y = y, yend = yend,
    x = x, xend = xend),
    size = 2, lineend = 'round', colour = "grey") +
  # median
  geom_segment(data = df.q2, aes(y = y, yend = yend,
    x = x, xend = xend),
    size = 0.5, lineend = 'round') +
  labs(x = "Values") + 
  # ggtitle("Random sample") +
  facet_grid(cols = vars(gp))
p
pA1 <- p
```

Thick vertical line = 3rd quartile.
Thin vertical line = 2nd quartile (median).

## Sample difference

### Illustrate results
```{r, fig.height=2}
nb <- 20

df <- tibble(bootid = 1,
             res = normdiff(q3(samp1), q3(samp2))
              )

p <- ggplot(df, aes(x = res, xend = res, y = bootid - 0.4, yend = bootid + 0.4)) + 
  theme_gar +
  geom_vline(xintercept = normdiff(q3(samp1),q3(samp2)),
             size = 2, colour = "grey") +
  geom_vline(xintercept = 0, size = 0.75, linetype = "dashed") +
  coord_cartesian(xlim = c(-0.4, 0.4)) +
  scale_x_continuous(breaks = seq(-0.5, 0.5, 0.1)) +
  labs(x = "Normalised difference") +
  theme(panel.grid.minor = element_blank(),
        axis.title.y = element_blank()) +
  ggtitle("(Group 1 - Group 2) / (Group 1 + Group 2)") +
  theme(plot.title = element_text(size = 16))
pA2 <- p
p
```
population normalised difference = `r round(normdiff(q3(samp1),q3(samp2)), digits = 2)`

Thick grey vertical line = difference between 3rd quartiles.
Vertical dashed line = zero reference line.

# Generate bootstrap samples & confidence intervals

We use 5,000 bootstrap samples.

```{r}
set.seed(666)
nboot <- 5000
bootsamp1 <- matrix(sample(samp1, nboot * n1, replace = TRUE), nrow = nboot)
bootsamp2 <- matrix(sample(samp2, nboot * n2, replace = TRUE), nrow = nboot)
```

## Illustrate 20 bootstrap samples

For each sample we superimpose the 3rd quartile---short vertical black line.
Disks = bootstrap observations.
Long vertical grey lines mark the sample 3rd quartiles.

```{r, fig.height=6}
nb <- 20
df <- tibble(res = c(as.vector(bootsamp1[1:nb,]),as.vector(bootsamp2[1:nb,])),
             bootid = c(rep(1:nb, each = n1),rep(1:nb, each = n2)),
             gp = factor(c(rep("Group 1",nb*n1),rep("Group 2",nb*n2))))

df2 <- tibble(bootid = rep(1:nb,2),
              res = c(apply(bootsamp1[1:nb,],1,q3),apply(bootsamp2[1:nb,],1,q3)),
              gp = factor(rep(c("Group 1", "Group 2"), each = nb))
              )

p <- ggplot(df, aes(y = bootid, x = res)) + theme_gar +
  # 3rd quartile
  geom_vline(data = df.q3, aes(group = gp, xintercept = x), size = 2, colour = "grey") +
  # scatterplots
  geom_point(alpha = 0.3, position = position_jitter(height=0.1)) +
  labs(x = "Values", y = "Bootstrap samples") +
  geom_segment(data = df2, aes(x = res, xend = res,
                               y = bootid - 0.4, yend = bootid + 0.4),
               size = 1.5) +
  theme(panel.grid.minor = element_blank()) +
  scale_y_continuous(breaks = seq(1, 20, 1), expand = expand_scale(mult = c(0.01, 0.01))) +
  facet_grid(cols = vars(gp))
pB1 <- p
p
```

## Illustrate bootstrap normalised differences

### Illustrate results
```{r, fig.height=6}
nb <- 20

df <- tibble(bootid = 1:nb,
             res = normdiff(apply(bootsamp1[1:nb,],1,q3), apply(bootsamp2[1:nb,],1,q3))
              )

p <- ggplot(df, aes(x = res, xend = res, y = bootid - 0.4, yend = bootid + 0.4)) + 
  theme_gar +
  geom_vline(xintercept = normdiff(q3(samp1),q3(samp2)), size = 2, colour = "grey") +
   geom_vline(xintercept = 0, size = 0.5, linetype = "longdash") +
  coord_cartesian(xlim = c(-0.4, 0.4)) +
  scale_x_continuous(breaks = seq(-0.5, 0.5, 0.1)) +
  labs(x = "Normalised differences", y = "Bootstrap samples") +
  geom_segment(size = 1.5) +
  theme(panel.grid.minor = element_blank()) +
  scale_y_continuous(breaks = seq(1, 20, 1), expand = expand_scale(mult = c(0.01, 0.01)))
pB2 <- p
p
```

# Bootstrap distributions

## Compute confidence intervals
```{r}
alpha <- 0.05
boot1.q3 <- apply(bootsamp1,1,q3)
boot2.q3 <- apply(bootsamp2,1,q3)
ci1 <- quantile(boot1.q3, probs = c(alpha/2, 1-alpha/2))
ci2 <- quantile(boot2.q3, probs = c(alpha/2, 1-alpha/2))
bootdiff <- normdiff(boot1.q3, boot2.q3)
ci.diff <- quantile(bootdiff, probs = c(alpha/2, 1-alpha/2)) 
```

## Bootstrap distributions of group quantiles
```{r, fig.height=4}

df <- tibble(x = c(boot1.q3, boot2.q3),
             gp = factor(c(rep("Group 1",nboot),rep("Group 2",nboot))))

# df.q3 <- tibble(q3 = c(q3(samp1), q3(samp2)),
#                  group = factor(c("Group 1", "Group 2")))

df.ci <- tibble(x = c(ci1[1], ci2[1]),
                xend = c(ci1[2], ci2[2]),
                y = c(0, 0),
                yend = c(0, 0),
                gp = factor(c("Group 1", "Group 2")))
  
p <- ggplot(df, aes(x = x)) + theme_gar +
  # sample q3: vertical line + label
  geom_vline(data = df.q3, aes(xintercept = x, group = gp), size = 2, colour = "grey") +
  # density
  geom_line(stat = "density", size = 1) +
  # confidence interval ----------------------
  geom_segment(data = df.ci,
               aes(x = x, xend = xend, y = y, yend = yend),
               lineend = "round", size = 2, colour = "black") +
  labs(x = "Bootstrap quantiles",
       y = "Density") +
  theme(axis.text.y=element_blank(),  #remove y axis labels
        axis.ticks.y=element_blank()) +
  facet_grid(cols = vars(gp))
p
pC1 <- p
```

## Group normalised differences
```{r, fig.height=4}
samp.nd <- normdiff(q3(samp1), q3(samp2))

df <- tibble(x = bootdiff)

p <- ggplot(df, aes(x = x)) + theme_gar +
  # sample 3rd quartile: vertical line
  geom_vline(xintercept = samp.nd, linetype = 'solid', size = 2, colour = "grey") +
      geom_line(stat = "density", size = 1) +
  labs(x = "Bootstrap differences",
       y = "Density") +
  theme(axis.text.y=element_blank(),  #remove y axis labels
        axis.ticks.y=element_blank()) +
  # confidence interval ----------------------
  geom_segment(x = ci.diff[1], xend = ci.diff[2],
               y = 0, yend = 0,
               lineend = "round", size = 2, colour = "black") +
  # zero reference line
  geom_vline(xintercept = 0, size = 0.5, linetype = "longdash") +
  ggtitle(paste0(round(samp.nd,digits = 3),
                 " [",
                 round(ci.diff[1],digits = 3),
                 ", ",
                 round(ci.diff[2],digits = 3)
                 ,"]"))
p
pC2 <- p
```

# Make summary figure
```{r, eval = FALSE}

pvoid <- ggplot() + theme_void() # empty plot for blank space

# pA <- cowplot::plot_grid(pA1, pA2,
#                     labels = NA,
#                     ncol = 2,
#                     nrow = 1,
#                     rel_widths = c(2, 1), 
#                     label_size = 20, 
#                     hjust = -0.5, 
#                     scale=.95)
# 
# pB <- cowplot::plot_grid(pB1, pB2,
#                     labels = NA,
#                     ncol = 2,
#                     nrow = 1,
#                     rel_widths = c(2, 1), 
#                     label_size = 20, 
#                     hjust = -0.5, 
#                     scale=.95)
# 
# pC <- cowplot::plot_grid(pC1, pC2,
#                     labels = NA,
#                     ncol = 2,
#                     nrow = 1,
#                     rel_widths = c(2, 1), 
#                     label_size = 20, 
#                     hjust = -0.5, 
#                     scale=.95)

p.LEFT <- cowplot::plot_grid(p.pop, pA1, pB1, pC1,
                    labels = c("A", "B", "C", "D"),
                    ncol = 1,
                    nrow = 4,
                    rel_heights = c(2, 1, 3, 2),
                    label_size = 20,
                    align = "v",
                    axis = "l",
                    hjust = -0.5,
                    scale=.95)

p.RIGHT <- cowplot::plot_grid(pvoid, pA2, pB2, pC2,
                    labels = NA,
                    ncol = 1,
                    nrow = 4,
                    rel_heights = c(2, 1, 3, 2),
                    label_size = 20,
                    align = "v",
                    axis = "l",
                    hjust = -0.5,
                    scale=.95)

cowplot::plot_grid(p.LEFT, p.RIGHT,
                    labels = NA,
                    ncol = 2,
                    nrow = 1,
                    rel_widths = c(2, 1),
                    label_size = 20,
                    align = "h",
                    axis = "b",
                    hjust = -0.5,
                    scale=.95)

# save figure
ggsave(filename=('./figures/figure_2indgps.pdf'),width=17,height=15)
ggsave(filename=('./figures/figure9.pdf'),width=17,height=15)
```

# Confidence interval with n = 100

```{r}
set.seed(21)
n1 <- 100
n2 <- 100
samp1 <- rlnorm(n1, meanlog = meanlog1, sdlog = sdlog1)
samp2 <- rlnorm(n2, meanlog = meanlog2, sdlog = sdlog2)
nboot <- 5000
bootsamp1 <- matrix(sample(samp1, nboot * n1, replace = TRUE), nrow = nboot)
bootsamp2 <- matrix(sample(samp2, nboot * n2, replace = TRUE), nrow = nboot)
alpha <- 0.05
boot1.q3 <- apply(bootsamp1,1,q3)
boot2.q3 <- apply(bootsamp2,1,q3)
# ci1 <- quantile(boot1.q3, probs = c(alpha/2, 1-alpha/2))
# ci2 <- quantile(boot2.q3, probs = c(alpha/2, 1-alpha/2))
bootdiff <- normdiff(boot1.q3, boot2.q3)
ci.diff <- quantile(bootdiff, probs = c(alpha/2, 1-alpha/2)) 
round(ci.diff, digits = 2)
```

# Confidence interval with n = 200

```{r}
set.seed(21)
n1 <- 200
n2 <- 200
samp1 <- rlnorm(n1, meanlog = meanlog1, sdlog = sdlog1)
samp2 <- rlnorm(n2, meanlog = meanlog2, sdlog = sdlog2)
nboot <- 5000
bootsamp1 <- matrix(sample(samp1, nboot * n1, replace = TRUE), nrow = nboot)
bootsamp2 <- matrix(sample(samp2, nboot * n2, replace = TRUE), nrow = nboot)
alpha <- 0.05
boot1.q3 <- apply(bootsamp1,1,q3)
boot2.q3 <- apply(bootsamp2,1,q3)
# ci1 <- quantile(boot1.q3, probs = c(alpha/2, 1-alpha/2))
# ci2 <- quantile(boot2.q3, probs = c(alpha/2, 1-alpha/2))
bootdiff <- normdiff(boot1.q3, boot2.q3)
ci.diff <- quantile(bootdiff, probs = c(alpha/2, 1-alpha/2)) 
round(ci.diff, digits = 2)
```

